---
title: "RCLabels"
author: "Matthew Kuperus Heun"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RCLabels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(RCLabels)
```

## Introduction

Working with matrices in the `matsbyname` world
often requires manipulating row and column labels 
to achieve desired outcomes for matrix manipulation.
The `RCLabels` package (row and column labels)
provides convenient tools for manipulating those labels.


## Use cases

Two applications of matrix mathematics are 
input-output analysis in economics and
the physical supply use table (PSUT) structure 
of energy conversion chain (ECC) analysis.
In those contexts, 
row and column labels describe processing stages or
flows of goods or services between processing stages.
Row and column labels can benefit those applications,
ensuring that like quantities are added, subtracted, multiplied, or divided, etc.,
provided that row and column labels are respected.

One package that respects row and column labels is `matsbyname`,
thereby making economic and ECC input-output analyses easier.
Easy manipulation of row and column labels
is an enabling capability.
This package (`RCLabels`) provides easy manipulation of row and column labels.
In fact, the `matsbyname` packages uses functions from `RCLabels` internally.


## Label structure

Row and column labels are always character strings and
often have a prefix--suffix structure, 
where the prefix and suffix are denoted by a separator
or delimited in other ways.
Example row and column labels include

* "pref -> suff" (separator "->")
* "pref [suff]" (suffix delimited by “ [” and “]”)
* "(pref) (suff)" (prefix and suffix both surrounded by “(” and “)”)

Prefixes are usually the "thing" of interest, e.g.
an energy carrier ("Coal") or 
a processing stage in an energy conversion chain
("Main activity producer electricity plants"). 
Suffixes are usually modifiers or metadata about the thing.
Suffixes can describe the destination of an energy carrier 
("Light [-> Industry in USA]").
Suffixes can describe the output of a processing stage 
("Production [of Coal in ZAR]"). 


## Working with row and column labels

The `RCLabels` package streamlines working with row and column labels.


### Notation

`RCLabels` enables creation of notation objects that describe the structure 
of a row or column label.

```{r}
# Create a notation object
my_notation <- notation_vec(pref_start = "(", pref_end = ") ",
                            suff_start = "[", suff_end = "]")
# Notation objects are character vectors
my_notation
```

Several notation objects are provided for convenience within RCLabels.

```{r}
arrow_notation
paren_notation
bracket_notation
from_notation
of_notation
to_notation
bracket_arrow_notation
```


### Creating row and column labels

Row and column labels can be created with the `paste_pref_suff()` function

```{r}
my_label <- paste_pref_suff(pref = "Coal", suff = "from Coal mines in USA", 
                            notation = my_notation)
my_label
```


### Manipulating row and column labels (prefixes and suffixes)

Row and column labels can be manipulated using several helpful functions.

```{r}
split_pref_suff(my_label, notation = my_notation)
flip_pref_suff(my_label, notation = my_notation)
switch_notation(my_label, from = my_notation, to = paren_notation)
switch_notation(my_label, from = my_notation, to = paren_notation, flip = TRUE)
```

The prefix or suffix can be extracted from a row or column label.

```{r}
keep_pref_suff(my_label, keep = "pref", notation = my_notation)
keep_pref_suff(my_label, keep = "suff", notation = my_notation)
```


### Vectors and lists of row and column labels

The functions in `RCLabels` work with vectors and lists of row and column labels.

```{r}
labels <- c("a [of b in c]", "d [of e in f]", "g [of h in i]")
split_pref_suff(labels, notation = bracket_notation)
```

This feature means that the functions in `RCLabels` can be used on data frames.
Note that `r `transpose = TRUE` ensures that a single list
column is created.

```{r}
df <- tibble::tibble(labels = labels)
result <- df |> 
  dplyr::mutate(
    splitted = split_pref_suff(labels, notation = bracket_notation, transpose = TRUE)
  )
result$splitted[[1]]
result$splitted[[2]]
result$splitted[[3]]
```


## Nouns and prepositions

As discussed above, the prefix is often the "thing" of interest, and
the remainder of the label (the suffix) modifies the prefix.
This use case is so common that we introduce additional terms 
that enable additional functionality.
The prefix is usually a _noun_ (one or more words), and
the suffix usually consists of _prepositional phrases_
(each consisting of a preposition and an object).
`RCLabels` includes a list of common prepositions.

```{r}
prepositions
```


## Working with row and column labels (nouns and prepositions)

`RCLabels` supports the "nouns and prepositions" view
of row and column labels 
with several functions.
`get_nouns()` extracts the nouns from a row or column label.

```{r}
get_nouns(labels, notation = bracket_notation)
```

`get_pps()` extracts prepositional phrases.

```{r}
get_pps(labels, notation = bracket_notation)
```

`get_prepositions()` extracts the prepositions themselves.

```{r}
get_prepositions(labels, notation = bracket_notation)
```

`get_objects()` returns the objects of those prepositions.
Note the the objects are named by the preposition of their phrase.

```{r}
get_objects(labels, notation = bracket_notation)
```

`split_labels()` breaks labels into nouns and objects with names
of "noun" and the prepositions.

```{r}
splitted_labels <- split_labels(labels, 
                                prepositions = prepositions, 
                                notation = bracket_notation)
splitted_labels
```

Row and column labels that have been split can be recombined.

```{r}
recombine_labels(splitted_labels, notation = bracket_notation)
recombine_labels(splitted_labels, notation = paren_notation)
```


## Modifying row and column labels

To modify row and column labels, use one of the `modify_*` functions.
`modify_nouns()` replaces nouns with other values.

```{r}
modify_nouns(labels, 
             new_nouns = c("Coal", "Oil", "Natural gas"), 
             notation = bracket_notation)
```

To modify other pieces of labels, use the `modify_label_pieces()` function.
`modify_label_pieces()` enables assigning new values using a "one-to-many" approach
that enables aggregation.

```{r}
modify_label_pieces(labels, 
                    piece = "noun", 
                    mod_map = list(Production = c("a", "b", "c", "d"),
                                   Manufacture = c("g", "h", "i", "j")), 
                    notation = bracket_notation)
modify_label_pieces(labels, 
                    piece = "in", 
                    mod_map = list(GHA = "c", ZAF = c("f", "i")), 
                    notation = bracket_notation)
modify_label_pieces(labels, 
                    piece = "of", 
                    mod_map = list(Coal = "b", `Crude oil` = c("e", "h")), 
                    notation = bracket_notation)
```

To eliminate a piece of a label altogether, use the `remove_label_pieces()` function.

```{r}
remove_label_pieces(labels, 
                    piece = "in", 
                    notation = bracket_notation)
remove_label_pieces(labels, 
                    piece = c("of", "in"), 
                    notation = bracket_notation)
```

With much power comes much responsibility!


## Conclusion

The `RCLabels` package streamlines the manipulation of row and column labels 
for matrices. 
Applications include input-output analysis in economics and energy conversion chain analysis
or anywhere row and column labels are important for matrix mathematics.
