---
title: "RCLabels"
author: "Matthew Kuperus Heun"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RCLabels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(RCLabels)
```

## Introduction

Working with matrices in the `matsbyname` world
often requires manipulating row and column labels 
to achieve desired outcomes for matrix manipulation.
The `RCLabels` package (row and column labels)
provides convenient tools for manipulating those labels.


## Label structure

Row and column labels are always character strings and
often have a prefix--suffix structure, 
where the prefix and suffix can be separated by a separator
or delimited in other ways.
Example labels include

* "pref -> suff" (separator "->")
* "pref [suff]" (suffix delimited by '' [" and "]")
* "(pref) (suff)" (prefix and suffix both surrounded by "(" and ")")

Prefixes are usually the "thing" of interest, e.g.
an energy carrier ("Coal") or 
a processing stage in an energy conversion chain
("Main activity producer electricity plants"). 
Suffixes are usually modifiers or metadata about the thing.
Suffixes can describe the destination of an energy carrier 
("Light [-> Industry in USA]") .
Suffixes can describe the output of a processing stage 
("Production [of Coal in ZAR]"). 


## Use cases

